---
title: AI coding assistance
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    toc: true
    toc-float: true
    page-layout: full
    number-sections: true
    number-depth: 3
    embed-resources: true
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    theme: [default, ../resources/tut-style.scss]
    css: ../resources/tut-style.css
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
bibliography: ../resources/references.bib
output_dir: "docs"
---

```{r setup, include=FALSE,warning=FALSE, cache=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE, message=FALSE, cache = TRUE, comment = "")
options(tinytex.engine = 'xelatex')
cleanRmdInput <- function(x) {
#x <- gsub("```\\{r","```markdown\n`r ''```\\{r",x)
x <- gsub("^```$","`` `",x)  # the Makefile will then change this back to ``` after pandoc
x
}
library(tidyverse)
library(pander)
FIG_PATH <- '14_ai_files/figure-html/'

# save the built-in output hook
hook_output <- knitr::knit_hooks$get("output")

# set a new output hook to truncate text output
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    more <- "..."
    n <- eval(parse(text=n))
    if (length(n) == 1) {
      if (length(x) > n) {
        # truncate the output
        x <- c(head(x, n), "....\n")
        }
    } else {
        x <- c(more, x[n], more)
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})
```

# Background information

AI code assistants are advanced tools powered by artificial
intelligence that help developers write, debug, and optimize code more
efficiently. These tools leverage machine learning models trained on
vast amounts of code to provide intelligent suggestions,
auto-completions, and even entire code snippets based on the context
of your work. They are designed to enhance productivity, reduce
repetitive tasks, and assist in learning new programming languages or
frameworks.

Some of the most popular AI code assistants include:

+ *GitHub Copilot*: Developed by GitHub and OpenAI, this tool
  integrates seamlessly with popular IDEs like Visual Studio Code,
  RStudio / Positron, Sublime Text, Vim and Emacs. It provides
  context-aware code suggestions and can even generate entire
  functions or classes.
+ *TabNine*: A general-purpose AI code assistant that supports
  multiple programming languages and integrates with various IDEs.
+ *Kite*: Focused on Python but also supports other languages, Kite
  offers intelligent code completions and documentation suggestions.
+ *Amazon CodeWhisperer*: A tool designed to assist developers in
  writing code faster by providing real-time suggestions and code
  snippets.
+ *OpenAI Codex*: The underlying model powering GitHub Copilot, Codex
  can be used directly via the OpenAI API for custom integrations.

AI code assistants are useful for several reasons:

+ *Increased Productivity*: They help automate repetitive coding
  tasks, allowing developers to focus on solving complex problems.
+ *Learning Aid*: They provide suggestions and examples, making it
  easier for beginners to learn new programming languages or
  frameworks.
+ *Error Reduction*: By suggesting syntactically correct code, they
  help reduce common errors and bugs.
+ *Code Optimization*: Some tools can suggest more efficient or
  idiomatic ways to write code.
+ *Documentation Assistance*: They can generate comments or
  documentation for your code, improving readability and
  maintainability.

## General Guidelines for Using AI Code Assistants

While AI code assistants are powerful tools, it is important to use
them effectively and responsibly. Here are some general guidelines:

1. *Understand the Suggestions*: Always review and understand the code
   suggestions provided by the AI. Do not blindly accept them, as they
   may not always align with your specific requirements or best
   practices.
2. *Use as a Learning Tool*: Treat the AI assistant as a mentor or
   guide, especially when learning a new language or
   framework. Analyze the suggestions to understand why they work.
3. *Maintain Security and Privacy*: Avoid sharing sensitive or
   proprietary information with the AI assistant, especially if it
   involves uploading code to external servers.
4. *Customize for Your Needs*: Many AI assistants allow customization
   to better suit your coding style or project requirements. Take
   advantage of these features to improve the relevance of
   suggestions.
5. *Stay Updated*: AI tools are constantly evolving. Keep your
   assistant updated to benefit from the latest features and
   improvements.
6. *Combine with Best Practices*: Use AI assistants as a supplement
   to, not a replacement for, good coding practices, thorough testing,
   and code reviews.
7. *Be Aware of Limitations*: AI assistants are not perfect and may
   occasionally provide incorrect or suboptimal suggestions. Use your
   judgment and expertise to evaluate their output.

By integrating AI code assistants like GitHub Copilot into your
workflow, you can significantly enhance your coding experience,
whether you're working in RStudio, Emacs, or any other development
environment.


# Github copilot

GitHub Copilot is an AI-powered code assistant developed by GitHub in
collaboration with OpenAI. It is designed to help developers write
code faster and more efficiently by providing intelligent code
suggestions, autocompletions, and even generating entire functions or
modules based on the context of the code. GitHub Copilot is powered by
OpenAI's Codex model, which is a descendant of the GPT (Generative
Pre-trained Transformer) family of models. Codex is specifically
trained on a large corpus of publicly available code from
repositories, documentation, and other programming-related resources.

GitHub Copilot supports a wide range of programming languages,
including R, Python, JavaScript, TypeScript, Go, Ruby, and more. It
integrates seamlessly with popular development environments such as
~Visual Studio Code~, ~RStudio~, ~JetBrains IDEs~, and ~Neovim~. The
tool is particularly useful for tasks such as writing boilerplate
code, exploring unfamiliar libraries, and improving productivity.

GitHub Copilot is available under different subscription plans,
including a free plan for students through the GitHub Student
Developer Pack.

## Obtaining a GitHub Student Developer Pack and Setting Up GitHub Copilot in RStudio

University students can access GitHub Copilot for free by obtaining a
GitHub Student Developer Pack. Below are the steps to acquire the pack
and set up GitHub Copilot in ~RStudio~:

1. **Sign Up for a GitHub Account**
   + If you don’t already have a GitHub account, go to
     <https://github.com> and sign up for a free account.

2. **Apply for the GitHub Student Developer Pack**
   + Visit the GitHub Student Developer Pack page:
     <https://education.github.com/pack>.
   + Click on the *Get your pack* button.
   + Verify your student status by providing proof of enrollment. This
     can include:
     - A valid school-issued email address.
     - A scanned copy of your student ID card.
     - Other official proof of enrollment (e.g., a transcript or
       enrollment letter).
   + Submit your application and wait for approval. This process may
     take a few days.

3. **Activate GitHub Copilot**
   + Once your GitHub Student Developer Pack is approved, go to the
     GitHub Copilot page: <https://github.com/features/copilot>.
   + Click on the *Start my free trial* button and follow the
     instructions to activate GitHub Copilot using your GitHub
     account.

4. **Install GitHub Copilot in RStudio** GitHub Copilot can be
   integrated into `RStudio` using the `Copilot for JetBrains`
   plugin. Follow these steps:
   + Open `RStudio`.
   + Navigate to ~Tools -> Add-ins -> Browse Add-ins~.
   + Search for the ~GitHub Copilot~ plugin and install it.
   + Once installed, log in to your GitHub account through the plugin
     to authenticate your access to GitHub Copilot.
   + After successful authentication, GitHub Copilot will be enabled
     in your ~RStudio~ environment.

5. **Start Using GitHub Copilot in RStudio**
   + Open an R script or create a new one in ~RStudio~.
   + Start typing your code, and GitHub Copilot will provide
     suggestions based on the context.
   + Use the ~Tab~ key to accept a suggestion or ~Esc~ to dismiss it.
   + You can also use comments to guide Copilot. For example, typing
     ~Create a function to calculate the mean of a vector~ will
     prompt Copilot to generate a relevant function.

## Final Notes

GitHub Copilot is a powerful tool that can significantly enhance your
coding experience in ~RStudio~. However, always review the suggestions
provided by Copilot to ensure they meet your specific requirements and
adhere to best practices. Additionally, be mindful of the ethical
considerations, such as avoiding the use of sensitive or proprietary
data in your code.


## Uses

Once copilot is activated, it is ready to start providing suggestions
inside your scripts.  If for example, you started a new R script and
type a `#` character, you might notice that along the lower right
border of the script window, there is a message like "Copilot: Waiting
for completions" or "Copilot: No completions available".  If you then
start to type out something like:

`# This script` and wait a second or so, copilot might make a
suggestion (which will appear as lighter coloured text).  If you
continue typing the suggestion will progressively be replaced by
suggestions that are consistent with what you are typing.  To accept
copilot's suggestion, you can hit the TAB key.

The same applies for when you are typing code.  After the first few
keystrokes, copilot will offer a suggestion.

Here is a list of the most common uses of this sort of AI assisted
coding:

1. Assisted auto-complete

   - context aware auto-complete as you type

     e.g. start typing:
     ```{r}
     #| label: ai test 1a
     #| results: markup
     #| eval: false
     #| echo: true
     #| cache: false
     calculate_mean <-
     ```

     Copilot will kick in an make a suggestion.  In my case that was:

     ```{r}
     #| label: ai test 1b
     #| results: markup
     #| eval: false
     #| echo: true
     #| cache: false
     calculate_mean <- function(x) {
       if (!is.numeric(x)) {
         stop("Input must be a numeric vector")
       }
       mean_value <- mean(x, na.rm = TRUE)
       return(mean_value)
     }
     ```

     ::: {.callout-note}
     Note that this suggestion even provides a safety net that ensures it
     fails elegantly when the input data is not numeric
     :::

     As another example, if we had just run the following (e.g. a
     linear model):

     ```{r}
     #| label: ai test 1c
     #| results: markup
     #| eval: false
     #| echo: true
     #| cache: false
     data_lm <- lm(mpg ~ wt + hp, data = mtcars)
     ```

     Copilot might make the following suggestion as an acknowledgement
     that this next step is commonly performed after fitting a linear
     model.

     ```{r}
     #| label: ai test 1d
     #| results: markup
     #| eval: false
     #| echo: true
     #| cache: false
     summary(data_lm)
     ```

2. Generating code / documentation from comments

   You can elicit suggestions by typing an instruction as a comment.
   For example, we could write a comment asking to fit a linear model
   of bill length against species from the penguins dataset:

   ```{r}
   #| label: ai test 2a
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   ## fit a linear model of bill length against species using the built in penguins dataset
   ```

   After I entered the above line in an empty R script and hit RETURN,
   copilot offered the following suggestion:

   ```{r}
   #| label: ai test 2b
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   data("penguins", package = "palmerpenguins")
   data_lm_penguins <- lm(bill_length_mm ~ species, data = penguins)
   ```

   ::: {.callout-note}

   Note, that it initially suggested the first
   line (starting with `data`) and once I accepted that line (by
   hitting TAB) the next line was suggested
   :::

   The instructive comments can span multiple lines and can include
   additional information that guides copilot to produce a code
   snippet that might satisfy your needs better.  For example, if we
   wanted to perform some data wrangling on the penguins data, we
   might stipulate that we want it to use tidyverse, the native R pipe
   and follow the style guidelines outlined by Hadley Wickham in R for
   Data Science

   ```{r}
   #| label: ai test 2c
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   ## calculate the mean and standard deviation bill length of each
   ## species of penguin using the tidyverse package, the native R
   ## (|>) pipe and follow the style guidelines outlined by Hadley
   ## Wickham in R for Data Science
   ```

   This resulted in the following:

   ```{r}
   #| label: ai test 2d
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   penguin_summary <- penguins |>
     group_by(species) |>
     summarize(
       mean_bill_length = mean(bill_length_mm, na.rm = TRUE),
       sd_bill_length = sd(bill_length_mm, na.rm = TRUE)
     )
   ```

    As another example, we could ask for a function for calculating
    the standard deviation of a numeric vector and to demonstrate its
    use on the penguin dataset

   ```{r}
   #| label: ai test 2e
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   ## generate a function for calculating the standard deviation of a
   ## numeric vector and to demonstrate its use on the penguin dataset
   ```

   Which resulted in:

   ```{r}
   #| label: ai test 2f
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   calculate_sd <- function(x) {
     if (!is.numeric(x)) {
       stop("Input must be a numeric vector")
     }
     sd_value <- sd(x, na.rm = TRUE)
     return(sd_value)
   }
   sd_bill_length <- calculate_sd(penguins$bill_length_mm)
   ```

   Perhaps I should have been more explicit that I wanted it to apply
   the function in the tidyverse manner!

3. Exploring libraries and APIs

   Copilot can help you explore and use R libraries by suggesting
   functions and their usage. This is particularly helpful when
   working with new or unfamiliar libraries.

   For example, if we start off typing `ggplot(`, copilot will provide
   us with the basic structure of a ggplot call from which we can
   learn about the function.

4. Refactoring code

   Copilot can suggest more efficient or cleaner ways to write your
   code, helping you refactor and improve its readability.

   Example: Refactoring a loop to use vectorized operations.

   ```{r}
   #| label: ai test 3a
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   # Calculate the square of each element in a vector
   x <- c(1, 2, 3, 4, 5)
   squares <- c()
   for (i in 1:length(x)) {
     squares[i] <- x[i]^2
   }
   ```

   If we then enter a comment like: `## can you refactor the above code using a more efficient approach?`

   If might offer a suggestion like:

   ```{r}
   #| label: ai test 3b
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   x <- c(1, 2, 3, 4, 5)
   squares_refactored <- x^2
   ```

   Similarly, we could have made a comment like: `## can you refactor
   the above code using the apply family of functions?`

   And it might have suggested:

   ```{r}
   #| label: ai test 3c
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   x <- c(1, 2, 3, 4, 5)
   squares_apply <- sapply(x, function(i) i^2)
   ```

5. Debugging assistance

   Copilot can help identify potential issues in your code and suggest
   fixes. For example, if you forget to handle missing values in a
   dataset, Copilot might suggest adding `na.rm = TRUE` to your function
   calls.

   ```{r}
   #| label: ai test 4a
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   # Calculate the mean of a column in a data frame
   calculate_column_mean <- function(df, column_name) {
     mean(df[[column_name]])
   }
   ```

   If we then ask it `## can you suggest ways to improve this function?`
   We might get:

   ```{r}
   #| label: ai test 4b
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   calculate_column_mean_improved <- function(df, column_name) {
     if (!is.data.frame(df)) {
       stop("Input must be a data frame")
     }
     if (!column_name %in% names(df)) {
       stop("Column not found in data frame")
     }
     if (!is.numeric(df[[column_name]])) {
       stop("Column must be numeric")
     }
     mean_value <- mean(df[[column_name]], na.rm = TRUE)
     return(mean_value)
   }
   ```

   ::: {.callout-note}

   Note that whilst this function is longer and more complex that our
   original function, it is arguably more robust.  Note also that
   copilot made the above suggestions line by line and I had to keep
   hitting TAB until it was complete.

   :::

6. Writing documentation for a function.


   GitHub Copilot can assist in generating *Roxygen* documentation for
   your R functions. Roxygen is a popular tool in the R ecosystem for
   documenting functions, datasets, and packages. It uses specially
   formatted comments to generate documentation files
   automatically. With Copilot, you can quickly create well-structured
   Roxygen comments, saving time and ensuring consistency in your
   documentation.

   This feature is particularly useful for creating detailed and
   standardized documentation for your functions, including
   descriptions, parameter explanations, and return values.

    Typically (and particularly if you are authoring an R package),
    functions should be documented so that users can read a help file
    that explains how the function works, what the parameters are and
    can try an example.  This documentation is structured in a format

    For example, we could ask it to `## generate Roxygen documentation
    for the above function`

   ```{r}
   #| label: ai test 6a
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   #' Calculate the mean of a specified column in a data frame
   #'
   #' This function calculates the mean of a specified numeric column in a given data frame,
   #' handling missing values appropriately.
   #'
   #' @param df A data frame containing the data.
   #' @param column_name A string specifying the name of the column for which to calculate the
   #' mean.
   #' @return The mean of the specified column, excluding any NA values.
   #' @examples
   #' df <- data.frame(a = c(1, 2, 3, NA
   #' ), b = c(4, 5, 6, 7))
   #' calculate_column_mean_improved(df, "a")
   #' @export
   #' @importFrom stats mean
   ```

6. Unit tests

   GitHub Copilot can assist in generating unit tests for your R
   functions, which is an essential practice for ensuring code
   reliability and correctness. Unit tests are small, isolated tests
   that verify the behavior of individual functions or components. In
   R, unit testing is commonly performed using the ~testthat~ package.

   With Copilot, you can quickly create test cases for your functions,
   including edge cases and typical use cases. This helps you save
   time and ensures that your code is robust and behaves as expected.

   Example: Generating unit tests for a function that calculates the
   mean of a numeric vector.

   If I now type `## generate unit tests for the calculate_mean function using the
   testthat package`:
   I might receive the following suggestion from copilot:

   ```{r}
   #| label: ai test 7a
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   library(testthat)
   test_that("calculate_mean works correctly", {
     expect_equal(calculate_mean(c(1, 2, 3, 4, 5)), 3)
     expect_equal(calculate_mean(c(1, 2, NA, 4, 5)), 3)
     expect_error(calculate_mean("not a numeric vector"), "Input must be a numeric vector")
   })
   ```

8. A chat agent


   GitHub Copilot can also function as a simple chat agent within your
   code editor. By starting a comment with `# q: `, you can ask
   Copilot a question, and it will attempt to provide an answer or
   explanation in the form of a comment starting with `# a: `. This
   feature is particularly useful for quick clarifications,
   understanding code, or getting guidance on how to implement
   specific functionality.

   This use case allows you to interact with Copilot conversationally,
   making it a helpful tool for learning, debugging, or brainstorming
   coding solutions.

   If I enter a comment like: `# q: explain the difference between a data frame and a tibble in R`

   Copilot might offer:

   ```{r}
   #| label: ai test 8a
   #| results: markup
   #| eval: false
   #| echo: true
   #| cache: false
   # a: A data frame is a base R data structure used for storing tabular data,
   #    while a tibble is a modern reimagining of the data frame provided by the
   #    tidyverse (specifically the tibble package). Tibbles have several advantages over
   #    traditional data frames, including better printing methods that show only the first
   #    few rows and columns, improved handling of variable types, and enhanced subsetting
   #    capabilities. Tibbles also do not convert character vectors to factors by default,
   #    which can help avoid unintended consequences during data analysis. Overall, tibbles
   #    are designed to be more user-friendly and consistent with tidyverse principles.
   ```


# Examples

The end
