---
title: Causal inference
author: "Murray Logan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
format:
  html:
    toc: true
    toc-float: true
    page-layout: full
    number-sections: true
    number-depth: 3
    embed-resources: true
    code-fold: false
    code-tools: true
    code-summary: "Show the code"
    code-line-numbers: true
    code-block-border-left: "#ccc"
    code-copy: true
    highlight-style: atom-one
    theme: [default, ../resources/tut-style.scss]
    css: ../resources/tut-style.css
crossref:
  fig-title: '**Figure**'
  fig-labels: arabic
  tbl-title: '**Table**'
  tbl-labels: arabic
engine: knitr
bibliography: ../resources/references.bib
output_dir: "docs"
---

```{r setup, include=FALSE,warning=FALSE, cache=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE, message=FALSE, cache = TRUE, comment = "")
options(tinytex.engine = 'xelatex')
cleanRmdInput <- function(x) {
                                        #x <- gsub("```\\{r","```markdown\n`r ''```\\{r",x)
  x <- gsub("^```$","`` `",x)  # the Makefile will then change this back to ``` after pandoc
  x
}
  library(tidyverse)
  library(pander)
  FIG_PATH <- '10_git_files/figure-html/'
```




::: {.newsbox}

**Other useful tutorials or resources**

- [Statistical Rethinking 2023 -05 - Elemental Confounds](https://www.youtube.com/watch?v=mBEA7PKDmiY&list=PLDcUM9US4XdPz-KxHM4XHt7uUVGWWVSus&index=6)
- [Applying the structural causal model framework for observational causal inference in ecology](https://esajournals.onlinelibrary.wiley.com/doi/10.1002/ecm.1554)
:::


# Preparations

This tutorial makes use of a couple of specialist R packages
(`dagitty` and `ggdag`) in addition to the `tidyverse` and Bayesian
modelling package (`brms`). So we will start by loading the necessary
libraries

```{r}
#| label: libraries
#| output: false
#| eval: true
#| warning: false
#| message: false
#| cache: false
library(tidyverse)
library(brms)
library(dagitty)
library(ggdag)
library(patchwork)
```


# Background

Understanding the relationships between variables is a fundamental
aspect of ecological research. However, interpreting these
relationships requires careful consideration of statistical principles
and study design. Carefully controlled experimental designs have
traditionally been the gold standard for inferring causation. However,
observational studies are becoming increasingly important to answer
large-scale ecological questions for which experiments are not
feasible or suitable. Whilst most studies aim to answer causal
questions, doing so can be challenging and requires careful
consideration of causal relationships if we are to avoid confounding
and other biases - hence the phrase "correlation does not imply
causation".

Moreover, there is often an urge to measure as many variables as
possible just in case they might turn out to be useful in statistical
analyses - particularly in observational studies, or surveys. Then
when it comes to performing the statistical analyses, there is a
temptation to put everything in the model under the false belief that
the model will be "better". Blindly doing so will often result in the
models behaving quite differently to what you might expect and produce
spurious and missleading estimates.

This tutorial introduces the concepts of correlation, confounding, and
causation, emphasizing their importance in ecological studies as well
as how we can identify and adjust models for causal inference.

# Structural Causal Modelling (SCM) framework

Structural Causal Modelling (SCM) is a useful and robust logic-based
framework for identifying cause and effect relationships and thus
supporting causal inference from observational data. The framework
comprises the following seven steps (the details each step will be
covered below):

1. Create a conceptual model of the system of interest

2. Choose a statistical test

3. Test the consistency of your conceptual model with the data

4. Identify biases (confounding, overcontrol, collider) that you need
   to adjust for when testing the cause of interest

5. Test for causality with the appropriate statistical test

6. Repeat steps 3-5 for each cause of interest

7. Repeat steps 1-6 for other conceptual models


## Step 1 - create the conceptual model

The conceptual model represents the researchers' understanding of the
system of interest as well as their assumptions about the causal
structure between variables. This structure should be based on domain
knowledge, prior research, and expert opinion.

For example, a conceptual model might indicate that `X` effects `Y`
(`X -> Y`), or perhaps `X` and `Y` might both be effected by `Z` and
thus `Z` is a confounder of the relationship between `X` and `Y` (`X
<- Z -> Y`).

The conceptual model may include both observed and unobserved
variables. Essentially, the conceptual model encodes our structural
assumptions about the relationships in the system. Importantly, they
do not impose the manner in which those relationships are
investigated. That is, they are agnostic to the kinds of statistical
models you use - they could be linear/additive/non-linear,
frequentist/Bayesian, hierarchical etc.

The conceptual model is visually represented as a **Directed Acyclic
Graph** (**DAG**) where nodes (dots) represent variables and arrows
represent assumed causal relationships. Following are some important
properties of a DAG:

- the direction of an arrow indicates the direction of causality, e.g.
  `X -> Y` means that `X` causes `Y`.
- it is _acyclic_, meaning that there are no cycles or loops in the
  graph, i.e. you cannot go from `X` to `Y` and then back to `X`.
- the presence of an arrow implies a possible relationship
- the absence of an arrow encodes an explicit assumption of no
  relationship

To help describe DAGs, I think it would be useful to provide some
examples with real variables. So lets consider a hypothetical example
presented in @Arif_2022 that explored the effects of marine protected
areas (MPAs) on reef fish biomass.


```{tikz}
%| label: dag_1
%| engine: tikz
%| echo: false
%| fig-width: 13
%| fig-height: 6
%| out-width: 500px
%| cache: true
%| class: tikz
%| engine-opts:
%|   template: "../resources/tikz-minimal.tex"
%\tikzstyle{HandLabel} = [font={\fontspec[Scale=1.1]{xkcd}}]
\tikzstyle{HandLabel} = [font={\fontspec[Scale=1.0]{ArchitectsDaughter-Regular}}]
\tikzstyle{Title} = [font={\fontspec[Scale=1.0]{ArchitectsDaughter-Regular}}]
%\tikzstyle{Title} = [font={\fontspec[Scale=0.8]{CompleteinHim}}]
\tikzstyle{Messy} = [decorate,decoration={random steps,segment length=3pt, amplitude=0.5pt}]
\begin{tikzpicture}
\node[Messy, HandLabel, draw, anchor=west] (mpa){MPA};
\node[Messy, HandLabel, draw, align=left, anchor=west] at ($(mpa.east) + (1cm, 0cm)$) (fishing){Fishing pressure};
\node[Messy, HandLabel, draw, align=left, anchor=west] at ($(fishing.east) + (1cm, 0cm)$) (biomass){Fish biomass};
\node[Messy, HandLabel, draw, align=left, anchor=south] at ($(mpa.north) + (0cm, 0.75cm)$) (depth){Depth};
\node[Messy, HandLabel, draw, align=left, anchor=south] at ($(fishing.north) + (0cm, 0.75cm)$) (complex){Structural complexity};
\node[Messy, HandLabel, draw, align=left, anchor=south] at ($(biomass.north) + (0cm, 0.75cm)$) (human){Human gravity};
\node[Messy, HandLabel, draw, align=left, anchor=north] at ($(fishing.south) + (0cm, -0.75cm)$) (coral){Coral cover};
\draw[-latex, Messy] (mpa.east) to (fishing);
\draw[-latex, Messy] ($(fishing.east) +(0cm, 0cm)$) to (biomass);
\draw[latex-, Messy] ($(mpa.north) +(0cm, 0cm)$) to (depth);
\draw[latex-, Messy] ($(fishing.north) +(0cm, 0cm)$) to (depth);
\draw[-latex, Messy] (complex) to (mpa);
\draw[-latex, Messy] (complex) to (biomass);
\draw[-latex, Messy] (human) to (fishing);
\draw[-latex, Messy] (human) to (biomass);
\draw[-latex, Messy] (mpa) to (coral);
\draw[-latex, Messy] (biomass) to (coral);
\end{tikzpicture}
```

There is an R package (**dagitty**) to help with the construction and
interrogation of DAGs. For constructing the DAG, we can use either the
`dagitty()` function or the `dagify()` function.

```{r}
#| label: scales
#| results: markup
#| eval: true
#| echo: true
#| cache: false
ls()
conflicts()
```

::: {.panel-tabset}
## `dagitty()`

```{r}
#| label: dag_1a
#| results: markup
#| eval: true
#| fig-width: 8
#| fig-height: 5
#| echo: true
#| cache: false
dag <- dagitty("
dag {
  depth -> mpa
  depth -> fishing
  mpa -> fishing
  fishing -> biomass
  complexity -> mpa
  complexity -> biomass
  human -> fishing
  human -> biomass
  mpa -> coral
  biomass -> coral
}
")
## ggdag(dag, text = TRUE, text_size = 2.5) +
##   theme_dag_blank()
```
